---
title: "Get started with cubature"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
output:
  html_document:
  fig_caption: yes
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Get started with cubature}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=FALSE}
knitr::opts_chunk$set(
    message = FALSE,
    warning = FALSE,
    error = FALSE,
    tidy = FALSE,
    cache = FALSE
)
```

## Introduction

`cubature` approximates definite integrals over hypercubes in arbitrary
dimension, including improper integrals over infinite limits, by
evaluating the integrand at adaptively-chosen sample points. It bundles
two well-known C libraries:

- the
  [cubature library](https://github.com/stevengj/cubature)
  of Steven G. Johnson (routines `hcubature` and `pcubature`), and
- the
  [Cuba library](https://feynarts.de/cuba/)
  of Thomas Hahn (routines `cuhre`, `divonne`, `suave`, `vegas`).

Together these give you six integration methods in a single package.
Each method is available through its own named function
(`hcubature`, `pcubature`, `cuhre`, `divonne`, `suave`, `vegas`) as
well as through a uniform dispatcher, `cubintegrate()`, that lets
you switch methods with a single argument. Vectorized variants of
every method are also available for performance-sensitive work; see
the [vectorization vignette](vectorization.html) for timing data
and recommendations.

## Your first integral

The simplest possible call integrates a one-argument function over
an interval.

```{r}
library(cubature)
hcubature(f = sin, lowerLimit = 0, upperLimit = pi)
```

The returned list has four elements:

- **`integral`** — the estimated value of the integral,
- **`error`** — the estimated absolute error,
- **`functionEvaluations`** — how many times the integrand was called,
- **`returnCode`** — an integer status code, where `0` means the
  algorithm reached the requested tolerance. A non-zero value means
  something went wrong; see `?hcubature` for the full meaning
  (`1` = hard internal failure, `2` = `maxEval` budget exhausted
  before tolerance was met).

Here is a two-dimensional example that integrates the product
`cos(x₁)·cos(x₂)` over the unit square.

```{r}
f <- function(x) cos(x[1]) * cos(x[2])
hcubature(f, lowerLimit = c(0, 0), upperLimit = c(1, 1))
```

The exact answer is $\sin(1)^2 \approx 0.7081$.

The integration limits can be infinite — `hcubature` and `pcubature`
automatically transform the variable using `atan` so the unit-hypercube
routines inside still work:

```{r}
## Standard normal density integrates to 1 over the real line
hcubature(function(x) exp(-x^2 / 2) / sqrt(2 * pi),
          lowerLimit = -Inf, upperLimit = Inf)
```

## The unified `cubintegrate()` interface

The package exposes a second, uniform way to call every integration
method: `cubintegrate()`. One signature, one argument convention, and
a `method = ` argument to pick which C routine you want. This is the
interface we recommend for new code, because swapping methods to
compare results or tune performance becomes a one-argument change.

```{r}
## The same 2D integral via three different methods, all the same way
cubintegrate(f, lower = c(0, 0), upper = c(1, 1), method = "hcubature")$integral
cubintegrate(f, lower = c(0, 0), upper = c(1, 1), method = "cuhre")$integral
cubintegrate(f, lower = c(0, 0), upper = c(1, 1), method = "pcubature")$integral
```

Each of those would otherwise have required `hcubature(...)`,
`cuhre(...)`, `pcubature(...)` — different function names, slightly
different argument names, and (for the Cuba routines) different
conventions for passing method-specific options. `cubintegrate`
papers over all of that.

Method-specific arguments (e.g. Cuhre's `key` parameter or Divonne's
`maxPass`) are passed through `...`. To see what each method
accepts, use `default_args()`:

```{r}
str(default_args(), list.len = 4, vec.len = 2)
```

## Choosing a method

The six methods are not interchangeable — each is best on a different
class of integrand. A short decision table:

| Situation | Recommended |
|---|---|
| You're not sure / general use | `hcubature` |
| Smooth integrand, low dimension (≤ 3) | `pcubature` or `cuhre` |
| Non-smooth, localized, or suspicious | `cuhre`, or `hcubature(..., robust = TRUE)` |
| High dimension (> 5) | `vegas`, `suave`, `divonne` (Monte Carlo) |
| You want a second opinion | any of the above, compared |

`hcubature` uses h-adaptive Genz–Malik subdivision; it refines the
integration region where the integrand varies most. `pcubature`
increases the polynomial degree of a tensor Clenshaw–Curtis rule on a
fixed domain; it can be very efficient on smooth integrands but is
brittle when features are localized (see the robustness section
below). Cuba's `cuhre` is h-adaptive like `hcubature` but with a
denser base rule, which makes it substantially more robust on
non-smooth integrands — at the cost of more function evaluations per
subregion. The Monte Carlo methods (`vegas`, `suave`, `divonne`)
scale better than deterministic methods in high dimensions but return
probabilistic error estimates rather than deterministic bounds.

When in genuine doubt, run your integrand through `cubintegrate` with
two different methods and check they agree.

## Vector-valued integrands

All six methods support integrands that return more than one scalar
per call — `fDim > 1` for `hcubature`/`pcubature`, `nComp > 1` for
the Cuba methods. The integrand should return a numeric vector of
length `fDim` (scalar interface) or a `fDim × npt` matrix (vectorized
interface). Example:

```{r}
## Simultaneously integrate sin and cos over [0, pi/2]
f2 <- function(x) c(sin(x), cos(x))
hcubature(f2, 0, pi / 2, fDim = 2)$integral
## Exact: both equal 1
```

## Robustness: when hcubature and pcubature can silently fail

The default error estimator in `hcubature` uses a single "null rule"
difference between the degree-7 Genz–Malik rule and its embedded
degree-5 companion on each region. For smooth integrands this is
reliable and cheap. For integrands with **localized support** — a thin
hinge, a near-step sigmoid, a very narrow Gaussian inside an oversized
bounding box — it can fail in a particular way: the rule's sample
points miss the integrand's support entirely, all sample values are
zero, the rule estimate is zero with zero estimated error, and the
algorithm confidently reports "converged" on a completely wrong
answer. This is distinct from ordinary numerical error — the algorithm
is internally consistent; it has just never evaluated the function at
a point where it's nonzero.  `pcubature` can fail the same way because
its Clenshaw–Curtis tensor grid is also a fixed sample geometry.

A small reproducible example: a hinge-squared times a steep logistic
times a standard Gaussian over `[-10, 10]²`. With default settings
`hcubature` returns exactly `0` (with `returnCode = 0`, i.e. silent
"success"), while the true integral is about 0.21:

```r
f <- function(x) {
  t1 <- -3.5862; t2 <- -9.3801
  beta0 <- 9.5;  gamma0 <- 19 * sqrt(0.75)
  (max(t1 + t2 * x[1] - x[2], 0))^2 *
    plogis(beta0 + gamma0 * x[1]) *
    dnorm(x[1]) * dnorm(x[2])
}
## Default: silently wrong
hcubature(f, c(-10, -10), c(10, 10))$integral
#> [1] 0
## robust = TRUE: correct
hcubature(f, c(-10, -10), c(10, 10), robust = TRUE)$integral
#> [1] ~0.21
## cuhre: also correct
cubintegrate(f, c(-10, -10), c(10, 10), method = "cuhre")$integral
#> [1] ~0.21
```

### The opt-in `robust` path

Starting in cubature 2.2.0, `hcubature()`, `pcubature()`, and
`cubintegrate()` accept an optional `robust = TRUE` argument that
activates extra safeguards. For `hcubature` these are: an additional
degree-3 null rule combined with the existing degree-5 one via a
Cuhre-style decay-check formula (Berntsen, Espelid & Genz, *ACM
TOMS* **17**(4), 437–451, 1991); a parent–child consistency check at
each subdivision step that inflates children's error estimates by
the discrepancy between their sum and the parent's prior estimate;
and a running peak-density "suspect region" detector. For
`pcubature` the robust path raises the initial Clenshaw–Curtis
tensor order so the starting grid is denser. In both cases the
default (non-robust) path is unchanged bit-for-bit; the safeguards
are paid for only when opted into.

The cost of `robust = TRUE` on smooth integrands is near zero in
most cases and at most ~25% extra function evaluations on
sharply-peaked ones. The cost on genuinely problem integrands (where
the default path was silently wrong) is several times more function
evaluations — which is the price of correctness. When `method` is a
Cuba method, `cubintegrate` silently ignores `robust` because those
integrators already use more conservative error estimators.

### When robust is not enough: the sampling-density cliff

Setting `robust = TRUE` is not a complete cure for either
`hcubature` or `pcubature`. The safeguards work on regions
where the rule has *seen some signal* and needs a better estimate of
it. If the integrand's support is so narrow that every sample point of
the base rule misses it entirely, the safeguards have nothing to work
with. This failure mode is determined by the rule's sample
geometry. For `hcubature`'s Genz–Malik rule on `[-h, h]^d`, the cliff
for hinge-line integrands sits at `t₁ = -√(9/70)·h ≈ -0.3586·h`; for
`pcubature`'s robust (m = 2) Clenshaw–Curtis grid it's at `t₁ =
-cos(3π/8)·h ≈ -0.3827·h`. For parameters on one side of the cliff
every sample point fires and the algorithm converges correctly; for
parameters on the other side every sample point evaluates to zero and
the algorithm silently reports zero. No combination of error-estimator
tweaks can detect this.

The only way to address the cliff is a **denser base rule**, which
is exactly what Cuba's Cuhre provides: it uses a degree-13 rule with
65 sample points in 2D, far more than Genz–Malik's 17, making the
cliff arbitrarily thin. **If your integrand is non-smooth and has
features much smaller than the integration box, prefer
`cubintegrate(method = "cuhre")` over `hcubature` / `pcubature` —
even with `robust = TRUE`.** As a secondary workaround, shrinking
the integration box (removing tail regions that contribute
negligibly) reduces the scale mismatch between the rule and the
integrand's features and can make either `hcubature` or `pcubature`
reliable on problems where they would otherwise fail.

### The `NOT_CONVERGED` return code

Independently of `robust`, starting in 2.2.0 `hcubature` and
`pcubature` return `returnCode = 2` (and emit an R `warning()`)
when the `maxEval` budget is exhausted without meeting the requested
tolerance. Previously such cases silently returned `returnCode = 0`
(success), with the caller expected to compare `error` against their
own tolerance. The new behavior is a strict improvement in honesty:
the `error` field is still populated with the best available
estimate, and the returned `integral` is still the best-effort
result, so users who want the permissive old behavior can wrap the
call in `suppressWarnings(...)` and/or check
`returnCode %in% c(0, 2)`. But callers that defensively did
`if (returnCode == 0) use(result)` will now correctly fall through
on non-convergence rather than silently accept a possibly-wrong
answer. The symbolic constants `CUBATURE_SUCCESS`, `CUBATURE_FAILURE`,
and `CUBATURE_NOT_CONVERGED` are exported from the upstream C header
for compiled clients.

## Where to go next

- The
  [vectorization vignette](vectorization.html) benchmarks scalar vs.
  vectorized interfaces across ten test integrands and makes
  per-method recommendations — read this if you call the integrator
  many times in a loop and want maximum performance.
- The reference pages for [`hcubature()`], [`pcubature()`],
  [`cubintegrate()`], [`cuhre()`], [`divonne()`], [`suave()`], and
  [`vegas()`] document every method-specific option.
- The source for both bundled C libraries is in
  `src/libcubature/` and `src/Cuba/`.

## Implementation notes

Both bundled C libraries are maintained as git submodules pointing at
personal forks with `R_pkg` branches that carry the R-package-specific
patches:

- The [cubature library fork](https://github.com/bnaras/libcubature/tree/R_pkg)
  (`src/libcubature`; the fork is named `libcubature` to avoid
  colliding with this R package's own repo name). The only
  substantive modification beyond the
  2.2.0 robustness safeguards is that we use `M = 16` rather than
  the default `M = 19` suggested by the original author for
  `pcubature`, to comply with CRAN package size limits.
- The [Cuba library fork](https://github.com/bnaras/Cuba/tree/R_pkg)
  (`src/Cuba`). Each time a new upstream release is made, we update
  the main branch and keep Unix-platform changes in the `R_pkg`
  branch. Windows customization lives in `Makevars.win`.

## Session info

```{r}
sessionInfo()
```
