---
title: "Vectorization benchmarks"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
output:
  html_document:
  fig_caption: yes
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Vectorization benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=FALSE}
knitr::opts_chunk$set(
    message = FALSE,
    warning = FALSE,
    error = FALSE,
    tidy = FALSE,
    cache = FALSE
)
```

## When to read this

This vignette assumes you've already read the
[Get started](cubature.html) vignette and you now have a specific
performance question: **is it worth vectorizing my R integrand?**

`cubature` exposes two calling conventions for the same underlying
C routine:

- **Scalar interface** (the default) — your R function receives one
  point at a time and returns one scalar (or length-`fDim` vector).
- **Vectorized interface** (opt-in, `vectorInterface = TRUE` for
  `hcubature`/`pcubature` or `nVec > 1` for Cuba methods) — your R
  function receives a matrix of points and returns a matrix of values,
  amortizing R's per-call overhead over many integrand evaluations at
  once.

The vectorized interface almost always wins for non-trivial problems,
sometimes by a factor of 10–50×, because the dominant cost is R's
function-call overhead rather than the arithmetic inside the
integrand. But the speedup depends on your specific integrand, the
method you pick, and the dimension. This vignette runs ten
representative integrands through every combination and reports
timings, so you can make an informed decision rather than guess.

## A timing harness

Our harness will provide timing results for `hcubature`, `pcubature`
(where appropriate), and Cuba `cuhre` calls — scalar and vectorized
variants of each.

```{r, message = FALSE}
library(bench)
library(cubature)

harness <- function(which = NULL,
                    f, fv, lowerLimit, upperLimit, tol = 1e-3, iterations = 20, ...) {

  fns <- c(hc = "Non-vectorized Hcubature",
           hc.v = "Vectorized Hcubature",
           pc = "Non-vectorized Pcubature",
           pc.v = "Vectorized Pcubature",
           cc = "Non-vectorized cubature::cuhre",
           cc_v = "Vectorized cubature::cuhre")
  cc <- function() cubature::cuhre(f = f,
                                   lowerLimit = lowerLimit, upperLimit = upperLimit,
                                   relTol = tol,
                                   ...)
  cc_v <- function() cubature::cuhre(f = fv,
                                     lowerLimit = lowerLimit, upperLimit = upperLimit,
                                     relTol = tol,
                                     nVec = 1024L,
                                     ...)
  hc <- function() cubature::hcubature(f = f,
                                       lowerLimit = lowerLimit,
                                       upperLimit = upperLimit,
                                       tol = tol,
                                       ...)
  hc.v <- function() cubature::hcubature(f = fv,
                                         lowerLimit = lowerLimit,
                                         upperLimit = upperLimit,
                                         tol = tol,
                                         vectorInterface = TRUE,
                                         ...)
  pc <- function() cubature::pcubature(f = f,
                                       lowerLimit = lowerLimit,
                                       upperLimit = upperLimit,
                                       tol = tol,
                                       ...)
  pc.v <- function() cubature::pcubature(f = fv,
                                         lowerLimit = lowerLimit,
                                         upperLimit = upperLimit,
                                         tol = tol,
                                         vectorInterface = TRUE,
                                         ...)

  ndim <- length(lowerLimit)

  if (is.null(which)) {
    fnIndices <- seq_along(fns)
  } else {
    fnIndices <- na.omit(match(which, names(fns)))
  }
  fnList <- lapply(names(fns)[fnIndices], function(x) call(x))

  argList <- c(fnList, iterations = iterations, check = FALSE)
  result <- do.call(bench::mark, args = argList)
  d <- result[seq_along(fnIndices), 1:9]
  d$expression <- fns[fnIndices]
  d
}
```

We reel off the timing runs.

## Example 1

```{r, message = FALSE}
func <- function(x) sin(x[1]) * cos(x[2]) * exp(x[3])
func.v <- function(x) {
    matrix(apply(x, 2, function(z) sin(z[1]) * cos(z[2]) * exp(z[3])), ncol = ncol(x))
}

d <- harness(f = func, fv = func.v,
             lowerLimit = rep(0, 3),
             upperLimit = rep(1, 3),
             tol = 1e-5,
             iterations = 100)[, 1:9]

knitr::kable(d, digits = 3, row.names = FALSE)
```


## Multivariate normal

Using `cubature`, we evaluate
$$
\int_R\phi(x)dx
$$
where $\phi(x)$ is the three-dimensional multivariate normal density with mean
0, and variance
$$
\Sigma = \left(\begin{array}{rrr}
1 &\frac{3}{5} &\frac{1}{3}\\
\frac{3}{5} &1 &\frac{11}{15}\\
\frac{1}{3} &\frac{11}{15} & 1
\end{array}
\right)
$$
and $R$ is $[-\frac{1}{2}, 1] \times [-\frac{1}{2}, 4] \times [-\frac{1}{2}, 2].$

We construct a scalar function (`my_dmvnorm`) and a vector analogue
(`my_dmvnorm_v`). First the functions.

```{r, message = FALSE}
m <- 3
sigma <- diag(3)
sigma[2,1] <- sigma[1, 2] <- 3/5 ; sigma[3,1] <- sigma[1, 3] <- 1/3
sigma[3,2] <- sigma[2, 3] <- 11/15
logdet <- sum(log(eigen(sigma, symmetric = TRUE, only.values = TRUE)$values))
my_dmvnorm <- function (x, mean, sigma, logdet) {
    x <- matrix(x, ncol = length(x))
    distval <- stats::mahalanobis(x, center = mean, cov = sigma)
    exp(-(3 * log(2 * pi) + logdet + distval)/2)
}

my_dmvnorm_v <- function (x, mean, sigma, logdet) {
    distval <- stats::mahalanobis(t(x), center = mean, cov = sigma)
    exp(matrix(-(3 * log(2 * pi) + logdet + distval)/2, ncol = ncol(x)))
}
```

Now the timing.

```{r, message = FALSE}
d <- harness(f = my_dmvnorm, fv = my_dmvnorm_v,
             lowerLimit = rep(-0.5, 3),
             upperLimit = c(1, 4, 2),
             tol = 1e-5,
             iterations = 10,
             mean = rep(0, m), sigma = sigma, logdet = logdet)
knitr::kable(d, digits = 3)
```

The effect of vectorization is huge. So it makes sense for users to
vectorize their integrands as much as possible for efficiency.

Furthermore, for this particular example, we expect `mvtnorm::pmvnorm`
to do pretty well since it is specialized for the multivariate normal.
The vectorized versions of `hcubature` and `pcubature` seem
competitive and in some cases better, as the table below shows.

```{r, message = FALSE}
library(mvtnorm)
g1 <- function() pmvnorm(lower = rep(-0.5, m),
                                  upper = c(1, 4, 2), mean = rep(0, m), corr = sigma,
                                  alg = Miwa(), abseps = 1e-5, releps = 1e-5)
g2 <- function() pmvnorm(lower = rep(-0.5, m),
                         upper = c(1, 4, 2), mean = rep(0, m), corr = sigma,
                         alg = GenzBretz(), abseps = 1e-5, releps = 1e-5)
g3 <- function() pmvnorm(lower = rep(-0.5, m),
                         upper = c(1, 4, 2), mean = rep(0, m), corr = sigma,
                         alg = TVPACK(), abseps = 1e-5, releps = 1e-5)

knitr::kable(bench::mark(g1(), g2(), g3(), iterations = 20, check = FALSE)[, 1:9],
             digits = 3, row.names = FALSE)
```

## Product of cosines

```{r, message = FALSE}
testFn0 <- function(x) prod(cos(x))
testFn0_v <- function(x) matrix(apply(x, 2, function(z) prod(cos(z))), ncol = ncol(x))

d <- harness(f = testFn0, fv = testFn0_v,
             lowerLimit = rep(0, 2), upperLimit = rep(1, 2), iterations = 1000)
knitr::kable(d, digits = 3)
```

## Gaussian function

```{r, message = FALSE}
testFn1 <- function(x) {
    val <- sum(((1 - x) / x)^2)
    scale <- prod((2 / sqrt(pi)) / x^2)
    exp(-val) * scale
}

testFn1_v <- function(x) {
    val <- matrix(apply(x, 2, function(z) sum(((1 - z) / z)^2)), ncol(x))
    scale <- matrix(apply(x, 2, function(z) prod((2 / sqrt(pi)) / z^2)), ncol(x))
    exp(-val) * scale
}

d <- harness(f = testFn1, fv = testFn1_v,
             lowerLimit = rep(0, 3), upperLimit = rep(1, 3), iterations = 10)

knitr::kable(d, digits = 3)
```

## Discontinuous function

```{r, message = FALSE}
testFn2 <- function(x) {
    radius <- 0.50124145262344534123412
    ifelse(sum(x * x) < radius * radius, 1, 0)
}

testFn2_v <- function(x) {
    radius <- 0.50124145262344534123412
    matrix(apply(x, 2, function(z) ifelse(sum(z * z) < radius * radius, 1, 0)), ncol = ncol(x))
}

d <- harness(which = c("hc", "hc.v", "cc", "cc_v"),
             f = testFn2, fv = testFn2_v,
             lowerLimit = rep(0, 2), upperLimit = rep(1, 2), iterations = 10)
knitr::kable(d, digits = 3)
```

## A simple polynomial (product of coordinates)

```{r, message = FALSE}
testFn3 <- function(x) prod(2 * x)
testFn3_v <- function(x) matrix(apply(x, 2, function(z) prod(2 * z)), ncol = ncol(x))

d <- harness(f = testFn3, fv = testFn3_v,
             lowerLimit = rep(0, 3), upperLimit = rep(1, 3), iterations = 50)
knitr::kable(d, digits = 3)
```

## Gaussian centered at $\frac{1}{2}$

```{r, message = FALSE}
testFn4 <- function(x) {
    a <- 0.1
    s <- sum((x - 0.5)^2)
    ((2 / sqrt(pi)) / (2. * a))^length(x) * exp (-s / (a * a))
}

testFn4_v <- function(x) {
    a <- 0.1
    r <- apply(x, 2, function(z) {
        s <- sum((z - 0.5)^2)
        ((2 / sqrt(pi)) / (2. * a))^length(z) * exp (-s / (a * a))
    })
    matrix(r, ncol = ncol(x))
}

d <- harness(f = testFn4, fv = testFn4_v,
             lowerLimit = rep(0, 2), upperLimit = rep(1, 2), iterations = 20)
knitr::kable(d, digits = 3)
```

## Double Gaussian

```{r, message = FALSE}
testFn5 <- function(x) {
    a <- 0.1
    s1 <- sum((x - 1 / 3)^2)
    s2 <- sum((x - 2 / 3)^2)
    0.5 * ((2 / sqrt(pi)) / (2. * a))^length(x) * (exp(-s1 / (a * a)) + exp(-s2 / (a * a)))
}
testFn5_v <- function(x) {
    a <- 0.1
    r <- apply(x, 2, function(z) {
        s1 <- sum((z - 1 / 3)^2)
        s2 <- sum((z - 2 / 3)^2)
        0.5 * ((2 / sqrt(pi)) / (2. * a))^length(z) * (exp(-s1 / (a * a)) + exp(-s2 / (a * a)))
    })
    matrix(r, ncol = ncol(x))
}

d <- harness(f = testFn5, fv = testFn5_v,
             lowerLimit = rep(0, 2), upperLimit = rep(1, 2), iterations = 20)
knitr::kable(d, digits = 3)
```

## Tsuda's example

```{r, message = FALSE}
testFn6 <- function(x) {
    a <- (1 + sqrt(10.0)) / 9.0
    prod( a / (a + 1) * ((a + 1) / (a + x))^2)
}

testFn6_v <- function(x) {
    a <- (1 + sqrt(10.0)) / 9.0
    r <- apply(x, 2, function(z) prod( a / (a + 1) * ((a + 1) / (a + z))^2))
    matrix(r, ncol = ncol(x))
}

d <- harness(f = testFn6, fv = testFn6_v,
             lowerLimit = rep(0, 3), upperLimit = rep(1, 3), iterations = 20)
knitr::kable(d, digits = 3)
```

## Morokoff & Caflisch example

```{r, message = FALSE}
testFn7 <- function(x) {
    n <- length(x)
    p <- 1/n
    (1 + p)^n * prod(x^p)
}
testFn7_v <- function(x) {
    matrix(apply(x, 2, function(z) {
        n <- length(z)
        p <- 1/n
        (1 + p)^n * prod(z^p)
    }), ncol = ncol(x))
}

d <- harness(f = testFn7, fv = testFn7_v,
             lowerLimit = rep(0, 3), upperLimit = rep(1, 3), iterations = 20)
knitr::kable(d, digits = 3)
```

## Wang-Landau sampling 1d, 2d examples

```{r, message = FALSE}
I.1d <- function(x) {
    sin(4 * x) *
        x * ((x * ( x * (x * x - 4) + 1) - 1))
}
I.1d_v <- function(x) {
    matrix(apply(x, 2, function(z)
        sin(4 * z) *
        z * ((z * ( z * (z * z - 4) + 1) - 1))),
        ncol = ncol(x))
}
d <- harness(f = I.1d, fv = I.1d_v,
             lowerLimit = -2, upperLimit = 2, iterations = 100)
knitr::kable(d, digits = 3)
```

```{r, message = FALSE}
I.2d <- function(x) {
    x1 <- x[1]; x2 <- x[2]
    sin(4 * x1 + 1) * cos(4 * x2) * x1 * (x1 * (x1 * x1)^2 - x2 * (x2 * x2 - x1) +2)
}
I.2d_v <- function(x) {
    matrix(apply(x, 2,
                 function(z) {
                     x1 <- z[1]; x2 <- z[2]
                     sin(4 * x1 + 1) * cos(4 * x2) * x1 * (x1 * (x1 * x1)^2 - x2 * (x2 * x2 - x1) +2)
                 }),
           ncol = ncol(x))
}
d <- harness(f = I.2d, fv = I.2d_v,
             lowerLimit = rep(-1, 2), upperLimit = rep(1, 2), iterations = 100)
knitr::kable(d, digits = 3)
```

## Summary

The recommendations are:

1. **Vectorize your integrand.** The time spent in so doing pays back
   enormously on all but the most trivial problems. This is easy to do
   and the examples above show how.

2. **Vectorized `hcubature` is a good default starting point.** It
   handles the widest range of problems well.

3. **For smooth integrands in low dimensions ($\leq 3$), try
   `pcubature`.** It can be substantially faster than `hcubature`
   when its assumptions hold. Experiment before committing in a
   production setting, and be aware of the sampling-density cliff
   documented in the Get Started vignette — when in doubt, cross-check
   with `cuhre`.

4. **For non-smooth or localized integrands, prefer
   `cubintegrate(method = "cuhre")`** or at least
   `hcubature(..., robust = TRUE)`. See the Robustness section of the
   [Get Started](cubature.html) vignette.

## Session info

```{r, message = FALSE}
sessionInfo()
```
